Method to infer relatedness using low coverage ancient DNA

### 2022-02-15 paper_overleaf/main.tex
Updates from Overleaf

\documentclass[12pt, letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage[super]{nth}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage[
singlelinecheck=false
]{caption}
\usepackage{amsmath}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{csvsimple}

@@

\newcommand{\BZ}{\mathbf{Z}}
\newcommand{\BD}{\mathbf{D}}
\newcommand{\BH}{\mathbf{H}}

@@

\begin{abstract}
\noindent Knowledge of familial relationships is an important part of several research fields. However, estimating relatedness from ancient DNA can be be difficult since DNA sequences extracted from ancient bones may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern humans. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and pairwise relatedness upto 3rd degree, while additionally differentiating between siblings and parent-child. We have developed another HMM to estimate ROH proportions, that can be utilized by the relatedness estimation model to output better estimates. 
\end{abstract}

@@

READ is a software that does not require population allele frequencies or a reference panel, and can work with around 0.5 x coverage in presence of ascertainment bias. The software solves the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudohaploids) are then compared to other individuals to calculate pairwise genetic distances, which in turn are used to infer relatedness. The limitations are that the software can identify upto only second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.

\subsection{How our method works}
Here, we present KIn (Kinship Inference), a Hidden Markov Model (HMM) to estimate relatedness from low-coverage ancient DNA data. KIn can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child. KIn is also able to incorporate ROH, contamination and ascertainment bias. We validate the performance of KIn using simulations and show that we are able to infer relatedness in an archaic human data set.

@@

\subsection{Relatedness Model overview} 
To estimate the relatedness of two individuals, we subdivide the genome into $L$ large windows $w$ (typically 10Mb). The inputs of our algorithm are i) the number of sites $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) position of ROH, by default obtained from ROH-HMM described in a following section.

@@

The model then uses this data for a given pair of individuals to classify each window into three hidden states $Z_w$ reflecting zero, one or two chromosomes shared IBD. Likewise, we classify each window into one of three possible ROH state $H_w$, reflecting, neither, one or both indivuals being in a run-of-homozygosity at this genomic location.

Ideally, we would like to compute the likelihood $P(\BD | \BZ, \BH, \theta$), where  $\theta$ is a vector of parameters (initial transition probability $\pi$, transition matrix $A$ and $N_w$).

@@

\begin{align}
P(\mathbf{D},\BZ, \BH|\theta) &= P(\mathbf{D}|\mathbf{Z},\BH, \theta) P(\BZ |\theta)P(\BH |\theta) \nonumber\\
&= \prod_w P(D_w|Z_w,H_w, \theta) \prod_w P(H_w |\theta) \prod_w P(Z_w |Z_{w-1},\theta) P(Z_0|\theta)
\end{align}

@@

Using the complete data likelihood allows us to factor the emission and transition probabilities, and hence we can maximize them independently.
Transition matrix for different cases of relatedness are obtained from simulations (see section). Initial transition probability $\pi_i$ is same for all the $i$.

\subsection{Emission probability}
In equation \ref{eq:cll}  the term $P(D_w | Z_w, H_w, \theta) = P(D_w | Z_w, H_w, N_w)$ is the emission probability, i.e. the probability of our data , given a particular hidden state. The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.

@@

If the two individuals are unrelated in a particular window and neither is in a ROH region (i.e. $Z_w = H_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes.

$p_0$, $p_1$, and $p_2$ are the expected proportion of differences in $i_0$, $i_1$, and $i_2$. $p_0$ is calculated directly from the data by calculating median of differences for all possible pairs of specimens. We calculate $p_1$ as $(3/4)p_1$ and $p_2$ as $p_1/2$. However, we need one more parameter (represented as $p_4$ in the table) in cases where IBD state is $i_1$ and ROH state is $\omega_2$, or IBD state is $i_2$ and ROH state is $\omega_1$ or $\omega_2$. In such cases the expected proportion of differences between the individuals should be 0. However, as mentioned before, we do all our calculations in windows of 10 Mb, and start/end positions of these windows may not coincide with that of ROH tracts. And so, the pairwise differences in this state (called $p_4$ in the table) can lie anywhere between 0 to $p_2$. We found $p_4$ = $p_2$/2 to be a reasonable assumption for expected pairwise proportion of differences in this IBD/ROH state.

Taken together, we can summarize $p$ as follows 
\begin{equation}\label{eq:p}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_1 & p_2 \\
p_0 & p_2 & p_4 \\
p_0 & p_4 & p_4
\end{array}\right]
\end{equation}

@@

We therefore add an overdispersion parameter $\delta$ that allows for variation in the $p$ between SNPs. This results in the betabinomial likelihood, (we show that such a distribution fits the data well (supplementary fig)).

\begin{equation}\label{eq:bbsimple}
P(D_{w}|p_i,\delta_i,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})}
\end{equation}

@@

In this step, using the standard forward-backward algorithm, we calculate the posterior probability of each IBD state in each window $\gamma_{wi}$. Expectation step, among other parameters, requires emission probabilities. We calculate emission probabilities for each case of $Z_w$ and $H_w$ from $\delta$ and $p$ using equation 3, and we get the total emission probabilities for each $Z_w$ as $\sum_{H_w} P(D_w|Z_w,N_w,H_w) P(H_w|\theta)$

@@

The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_j$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_wi$ and the ROH state-probabilities and $h_{w\omega}$.

\begin{equation}
\mathcal{C} = \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta, p)h_{w\omega}\gamma_{wi}
\end{equation}

Using equation \ref{eq:p}, we simplify this by grouping all the terms that would result in the same $p$, i.e.

\begin{align*}
k_{w0} &= \gamma_{w0}\\
k_{w1} &= \gamma_{w1} h_{w0}\\
k_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
k_{w4} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
\end{align*}

So we can rewrite Eqn 4 as:
\begin{align}
\mathcal{C} &= \sum_{w=0}^L\sum_{j=0}^4  \log P(D_{w}|Z_w, H_w, N_w, \delta_j, p_j)k_{wj}
\end{align}

Since each term in this equation depends on only one $\delta_j$, we can optimize for them independently.

We observed that estimating the $\delta_j$ without constraints can result in wide Beta distributions with very high variance, leading to over fitting of data. For example, a parent  shares exactly one chromosome IBD with its child throughout the genome, so we expect only one IBD state. However,  running an unconstrained model could use this data to fit a pair of siblings with a very wide beta-distribution (with a very high $\delta_1$, which causes problems in differentiating between parent-child and siblings.

@@

We choose t for each beta distribution to be proportional to the mean ($p$), since the variances of beta distribution corresponding to different $x$ states follow the same relation as the distribution means (supplement figure). 
\begin{align}
    Var(X_{x_1}) = 3/4 Var(X_{x_0})\\
    Var(X_{x_2}) = 1/2 Var(X_{x_0})
\end{align}

@@

\subsection{Emissions}
Similar to our relatedness model, we estimate emissions using EM algorithm with expectation and maximization steps. The expectation step outputs the posterior probability $\Gamma$ of being in $y_0$ and $y_1$ in each window. Maximization step uses the $\Gamma$ to optimize the $\Delta$ and $\rho$ parameters with a betabinomial cost function:

@@

In case of very noisy data, we see high pairwise differences in some windows that are not explained with the beta binomial distributions estimated for $y_0$ and $y_1$. To avoid our distributions from flattening to explain these high differences, we force these high difference windows to $y_1$ (see supplement figure).

@@

Contamination by present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020} . Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with  low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations (supplementary figure with READ?, lcMLkin at 4x). To address this issue, we  adjust the  pairwise differences $D_w$, given the contamination estimates of both samples and the average divergence between the target population and a putative contaminating population.

For each pair of individuals $i$ and $j$ we approximate the contamination $C_{i,j}$ as the sum of contamination for each individual $C_i + C_j$. Then, $C_{i,j}$ represents the probability of comparing a contaminant read from an individual to an endogenous read from the other individual. Throughout, we assume that contamination levels are low, so that we can ignore comparison between contaminant reads, i.e. $C_iC_j \approx 0$.  We denote the average divergence between target and contaminating populations as $p_c$, and the observed proportion of differences between individuals $i$ and $j$ as $p_{aij}$. We calculate average endogenous pairwise difference $p_eij$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
\begin{align}
    p_{aij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
    or, p_{eij} = \frac{p_{aij} - C_{ij}  p_c}{1-C_{ij}} 
\end{align}

@@

\paragraph{Short version}
With contamination, we do not observe the number of differences $D$ and total number of comparisons directly $N$, but instead need to approximate them. The window-subscript is omitted in the following for clarity:
\begin{equation}
    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) \approx 
    P( \hat{D_w} | \hat{N_w}, Z, \theta)
\end{equation}
We do that by calculating the probability that a single read is endogenous given it has a difference.
\begin{align}
    \hat{D} = \mathbb{E}[D | D_{\text{obs}}, c] &= 
    D_{\text{obs}} P(E_i | D_i=1) \\
    &= D_\text{obs} \frac{P(D_i=1 | E_i)P(E_i)}{P(D_i=1)}\\
    &= D_\text{obs} \frac{(1-c)p_e}{(1-c)p_e + c p_c}
\end{align}

@@

\begin{itemize}
\item W: Total number of windows
\item w: Index of window
\item $I$:IBD states base
\item $Z_w$: IBD state in window w
\item $\mathbf{Z}$: Vector of $Z_w$'s
\item $\omega$:ROH states base
\item $H_{w\omega}$: ROH state in $w$
\item $s$: SNP
\item $d_s$: pairwise difference at $s$
\item $\mathbf{R_{si}}$: Vector of reads mapped to snp s for individual $i$
\item $n_{si}$: Number of reads at SNP s for individual $i$
\item $a_{si}$: Number of reads with alternate allele at SNP s for individual $i$
\item $N_{wij}$: Number of SNPs overlapping for individuals $i$,$j$ in window $w$
\item $D_{w}$: Number of SNPs in window w where a pair of individuals differ
\item $\mathbf{D}$: Vector of $D_w$
\item $\pi$: Initial transition probability
\item $A$: Transition matrix
\item $theta$: vector of additional parameters ($\pi$, $A$ and $N_w$)
\item $\psi$: Matrix of parameters representing expected proportion of differences in different $I$ and $\omega$
\item $psi_{I\omega}$: Element on row $I$ and column $\omega$ of $\psi$ matrix
\item $x$: All combinations of $I$ and $\omega$ states with unique $\psi_{I\omega}$
\item $p$: Proportion of differences in $x$
\item $\alpha,\beta$: Parameters of Beta distribution
\item $\delta$: $\alpha + \beta$ for relatedness model
\item $\gamma_{wI}$: posterior probability of $I$ in $w$
\item $k_{wI}$: Posterior probability of $x$ in $w$ 
\item $\nu_{w}$: Allele frequency in window w
\item X: Random variable
\item r: Relatedness between individuals
\item $\theta$: Vector of initial probability of being in an IBD state $I_l$
\item $/mathbf{B}$: Beta function
\item $B$: Reparameterized Beta function
\item $Delta$: Heterozygosity in an individual
\item Y: Hidden Homozygosity state
\item $\Theta$: vector of additional parameters ($\pi$ $N_w$)
\item $Gamma_{wY}$: Posterior probability of Y in w
\item $/eta$: $/alpha + /beta$ for ROH model
\item $c$: Contamination estimate for a library
\item $p_c$: Average proportion of differences between endogenous and contaminating reads
\item $p_e$:Average proportion of differences between endogenous reads
\item $p_a$: Average observed proportion of differences
\item $D_{cor,w}$ Contamination corrected number of differences in $w$
\item $N_{cor,w}$ Contamination corrected number of overlapping sites in $w$
\item $\Xi$: Endogenous read
\item $S_{cor,w}$ $N_{cor,w}$ - $D_{cor,w}$
\item $/mathbf{D_cor}$ Vector of $D_{cor,w}$ 
\item E: Expectation
\item P: Probability
\end{itemize}

@@

We predict the following categories of relatedness with KIn: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated (including higher than third degree relatives). We show the working of Relatedness HMM with fig.1 showing performance of each model, when applied to a pair of siblings.

\begin{figure}[h]
    \centering
    \includegraphics[width=18cm]{plots/plotimg/IBDplot.png}
    \caption{Comparison of IBD fragments predicted by different relatedness models. Top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings. Dashed lines in this panel represent $p_0$, $p_1$ and $p_2$ estimates. The panel just below shows the true IBD state for each window. Each panel below shows the IBD states predicted by a particular relatedness model. The log likelihood value for each model is shown on upper left corner of the panel. }
    \label{fig:ibd}
\end{figure}

We see that in this case, the siblings model predictions match true IBD states the most. This is reflected in the log likelihoods values (LL) of the models as well, where siblings model has the highest value. We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48). All the models in fig.1 utilize information about ROH given by the ROH-HMM.

Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in fig 2. We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.1x.

\begin{figure}[h]
    \centering
    \includegraphics[width=18cm]{plots/inkscape_finalImg/ROHplot.png}
    \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment. Top panel shows proportion of differences in a simulated individual along the genome, and the bottom panel shows probability of seeing no ROH. (A) Simulation with no ascertainment, contamination, or ROH. (B) Simulation with ROH. (C) Simulation with ROH and ascertainment. (D) Simulation with ROH and contamination.}
    \label{fig:galaxy}
\end{figure}

>>>>

\documentclass[12pt, letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage[super]{nth}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage[
singlelinecheck=false
]{caption}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{csvsimple}

@@

\newcommand{\BZ}{\mathbf{Z}}
\newcommand{\BD}{\mathbf{D}}
\newcommand{\BH}{\mathbf{H}}
\newcommand{\Btheta}{\pmb{\theta}}

@@

\begin{abstract}
\noindent Knowledge of familial relationships is an important part of several research fields. However, estimating relatedness from ancient DNA can be be difficult since DNA sequences extracted from ancient bones may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern humans. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness upto 3rd degree, while additionally differentiating between siblings and parent-child. We have developed another HMM to estimate ROH proportions, that can be utilized by the relatedness estimation model to output better estimates. 
\end{abstract}

@@

READ is a software that does not require population allele frequencies or a reference panel, and can work with around 0.5 x coverage in presence of ascertainment bias. The software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudohaploids) are then compared to other individuals to calculate pairwise genetic distances, which in turn are used to infer relatedness. The limitations are that the software can identify upto only second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.

\subsection{How our method works}
Here, we present KIn (Kinship Inference), a Hidden Markov Model (HMM) to estimate relatedness from low-coverage ancient DNA data. KIn can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child. KIn is also able to incorporate ROH, contamination and ascertainment bias. We validate the performance of KIn using simulations and show that we are able to infer relatedness in a neandertal data set, and a Bronze age dataset.

@@

\subsection{Relatedness Model overview} 
To estimate the relatedness of two individuals, we subdivide the genome into $L$ large windows $w$ (typically of size 10Mb). The inputs of our algorithm are i) the number of sites $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) probability of ROH in in windows, by default obtained from ROH-HMM described in a following section. Calculation of $N_w$ and $D_w$ requires comparison of reads at all positions for a pair of individuals. For low coverage genomes, this can be done by comparing randomly sampled read at each position for a pair of individuals (or by comparing pseudo haploid sequences). However, we realized that such random sampling can throw away a lot of useful data. To avoid this wastage, our input generation pipeline for Relatedness HMM calculates proportion of reads with alternate allele, at all positions for each genome. We use these proportions to find pairwise heterozygosity as the probability of comparing different alleles. These probabilities are summed up in each window to give $D_w$, along with the number of overlapping sites $N_w$.

@@

The model then uses this data for a given pair of individuals to classify each window into three hidden states $Z_w \in$ ($i_0$, $i_1$, $i_2$), reflecting zero, one or two chromosomes shared IBD. Likewise, we classify each window into one of three possible ROH state $H_w \in$ ($\omega_0$, $\omega_1$, $\omega_2$), reflecting, neither, one or both individuals being in a run-of-homozygosity at this genomic location.

Ideally, we would like to compute the likelihood $P(\BD | \BZ, \BH, \theta)$, where $\theta$ is a vector of parameters (initial transition probability $\pi$, transition matrix $A$ and $N_w$).

@@

\begin{align}
P(\mathbf{D},\BZ, \BH|\theta) &= P(\mathbf{D}|\mathbf{Z},\BH, \theta) P(\BZ |\theta)P(\BH |\theta) \nonumber\\
&= \prod_w P(D_w|Z_w,H_w, \theta) \prod_w P(H_w |\theta) \prod_w P(Z_w |Z_{w-1},\theta) P(Z_0|\theta)
\label{eq:1}
\end{align}

@@

Using the complete data likelihood allows us to factor the emission and transition probabilities, and hence we can maximize them independently.
Transition matrix for different cases of relatedness are either derived from theory, or obtained from simulations (see section). Initial transition probability $\pi_i$ is same for all the $i$.

\subsection{Emission probability}
In equation \ref{eq:1} the term $P(D_w | Z_w, H_w, \theta) = P(D_w | Z_w, H_w, N_w)$ is the emission probability, i.e. the probability of our data , given a particular hidden state. The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.

@@

If the two individuals are unrelated in a particular window and neither is in a ROH region (i.e. $Z_w = i_0$, and $H_w = \omega_0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes.

$p_0$, $p_1$, and $p_2$ are the expected proportion of differences in $i_0$, $i_1$, and $i_2$. $p_0$ is calculated directly from the data by calculating median of differences for all possible pairs of specimens. We calculate $p_1$ as $(3/4)p_1$ and $p_2$ as $p_1/2$. However, we need one more parameter (represented as $p_3$ in the matrix below) in cases where IBD state is $i_1$ and ROH state is $\omega_2$, or IBD state is $i_2$ and ROH state is $\omega_1$ or $\omega_2$. In such cases the expected proportion of differences between the individuals should be 0. However, as mentioned before, we do all our calculations in large windows, and start/end positions of these windows may not coincide with that of ROH tracts. And so, the pairwise differences in this state (called $p_3$ in the table) can lie anywhere between 0 to $p_2$. We found $p_3$ = $p_2$/2 to be a reasonable assumption for expected pairwise proportion of differences in this IBD/ROH state.

Taken together, we can summarize $p$ as follows 
\begin{equation}\label{eq:2}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_1 & p_2 \\
p_0 & p_2 & p_3 \\
p_0 & p_4 & p_3
\end{array}\right]
\end{equation}

@@

We therefore add an overdispersion parameter $\delta$ that allows for variation in the $p$ between SNPs. This results in the betabinomial likelihood, (we show that such a distribution fits the data well (fig. S\ref{FigS1:binom})).

\begin{equation}\label{eq:3}
P(D_{w}|p_j,\delta_j,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_j \delta_{j}, N_w-D_w+ \delta_{j}(1-p_{j}))}{ B(p_{j}\delta_{j}, (1-p_{j})\delta_{j})}
\end{equation}

@@

In this step, using the standard forward-backward algorithm, we calculate the posterior probability of each IBD state in each window ($\gamma_{wi}$). Expectation step, among other parameters, requires emission probabilities. We calculate emission probabilities for each case of $Z_w$ and $H_w$ from $\delta$ and $p$ using equation \ref{eq:3}, and we get the total emission probabilities for each $Z_w$ as $\sum_{H_w} P(D_w|Z_w,N_w,H_w) P(H_w|\theta)$

@@

The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_j$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities and $h_{w\omega}$.

\begin{equation}\label{eq:4}
\mathcal{C} = \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta, p)h_{w\omega}\gamma_{wi}
\end{equation}

Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p$, i.e.

\begin{align*}
k_{w0} &= \gamma_{w0}\\
k_{w1} &= \gamma_{w1} h_{w0}\\
k_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
k_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
\end{align*}

So we can rewrite equation \ref{eq:4} as:
\begin{align}
\mathcal{C} &= \sum_{w=1}^L\sum_{j=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_j, p_j)k_{wj}
\end{align}

Since each term in this equation depends on only one $\delta_j$, we can optimize for them independently. For convenience, we refer to these four states corresponding to different $p$ as $x$ states.

We observed that estimating the $\delta_j$ without constraints can result in wide Beta distributions with very high variance, leading to over fitting of data (fig. \ref{figS4:bndsbeta}), and resulting in incorrect posterior probabilities for $Z_w$ (fig. \ref{figS3:bnds}). For example, a parent shares exactly one chromosome IBD with its child throughout the genome, so we expect only one IBD state. However, running an unconstrained model could use this data to fit a pair of siblings with a very wide beta-distribution (with a very high $\delta_1$, which may result in problems in differentiating between parent-child and siblings.

@@

We choose t for each beta distribution to be proportional to the mean ($p$), since the variances of beta distribution corresponding to different $x$ states follow the same relation as the distribution means. 
\begin{align}
    Var(X_{x_1}) = 3/4 Var(X_{x_0})\\
    Var(X_{x_2}) = 1/2 Var(X_{x_0})
\end{align}

Even though this may not hold true when we bin sites in large windows, we still find that $Var(X_{x_0}) > Var(X_{x_1}) > Var(X_{x_2})$ (fig. \ref{FigS1:binom}).

@@

\subsection{Emissions}
Similar to our relatedness model, we estimate emissions using EM algorithm with expectation and maximization steps. The expectation step outputs the posterior probability $\Gamma$ of being in $y_0$ and $y_1$ in each window. Maximization step uses the $\Gamma$ to optimize overdispersion parameter $\eta$ with a betabinomial cost function:

@@

In case of very noisy data, we see high pairwise differences in some windows that are not explained with the betabinomial distributions estimated for $y_0$ and $y_1$. To avoid our distributions from flattening to explain these high differences, we force these high difference windows to $y_1$ (fig. S\ref{figS5:ROHforced}).

@@

Contamination by present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020} . Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with  low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations. To address this issue, we  adjust $D_w$ and $N_w$, given the contamination estimates of both samples and the average divergence between the target population and a putative contaminating population.

For each pair of individuals $i$ and $j$ we approximate the contamination $C_{i,j}$ as the sum of contamination for each individual $C_i + C_j$. Then, $C_{i,j}$ represents the probability of comparing a contaminant read from an individual to an endogenous read from the other individual. Throughout, we assume that contamination levels are low, so that we can ignore comparison between contaminant reads, i.e. $C_iC_j \approx 0$.  We denote the average divergence between target and contaminating populations as $p_c$, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$. We calculate average endogenous pairwise difference $p_{eij}$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
\begin{align}
    p_{oij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
    or, p_{eij} = \frac{p_{oij} - C_{ij}  p_c}{1-C_{ij}} 
\end{align}

@@

\paragraph{Short version}
With contamination, we do not observe the number of differences $D_{true}$ and total number of comparisons $N_{true}$ directly, but instead need to approximate them. The window-subscript is omitted in the following for clarity:
\begin{equation}
    P(D_{\text{obs}} | N_\text{obs}, Z, \theta) \approx 
    P( \hat{D} | \hat{N}, Z, \theta)
\end{equation}
We do that by calculating the probability that a single read is endogenous given it has a difference.
\begin{align}
    \hat{D} = \mathbb{E}[D | D_{\text{obs}}, c] &= 
    D_{\text{obs}} P(E_i | D_i=1) \\
    &= D_\text{obs} \frac{P(D_i=1 | E_i)P(E_i)}{P(D_i=1)}\\
    &= D_\text{obs} \frac{(1-c)p_e}{(1-c)p_e + c p_c}
\end{align}

@@

\begin{itemize}
\item L: Total number of windows
\item w: Index of window
\item $N_w$: Number of sites in w
\item $D_{w}$: Number of pairwise differences in w
\item $Z_w$: Hidden IBD state in window w
\item $H_w$: ROH state in window w
\item A: Transition matrix
\item $\pi$: Vector of initial transition probabilities
\item $\theta$: Vector of $\pi$, A, $N_w$
\item $\mathbf{Z}$: Vector of $Z_w$'s
\item $\omega$:ROH states base
\item $H_{w\omega}$: ROH state in $w$
\item $s$: SNP
\item $d_s$: pairwise difference at $s$
\item $\mathbf{R_{si}}$: Vector of reads mapped to snp s for individual $i$
\item $n_{si}$: Number of reads at SNP s for individual $i$
\item $a_{si}$: Number of reads with alternate allele at SNP s for individual $i$
\item $N_{wij}$: Number of SNPs overlapping for individuals $i$,$j$ in window $w$
\item $D_{w}$: Number of SNPs in window w where a pair of individuals differ
\item $\mathbf{D}$: Vector of $D_w$
\item $\pi$: Initial transition probability
\item $A$: Transition matrix
\item $theta$: vector of additional parameters ($\pi$, $A$ and $N_w$)
\item $\psi$: Matrix of parameters representing expected proportion of differences in different $I$ and $\omega$
\item $psi_{I\omega}$: Element on row $I$ and column $\omega$ of $\psi$ matrix
\item $x$: All combinations of $I$ and $\omega$ states with unique $\psi_{I\omega}$
\item $p$: Proportion of differences in $x$
\item $\alpha,\beta$: Parameters of Beta distribution
\item $\delta$: $\alpha + \beta$ for relatedness model
\item $\gamma_{wI}$: posterior probability of $I$ in $w$
\item $k_{wI}$: Posterior probability of $x$ in $w$ 
\item $\nu_{w}$: Allele frequency in window w
\item X: Random variable
\item r: Relatedness between individuals
\item $\theta$: Vector of initial probability of being in an IBD state $I_l$
\item $/mathbf{B}$: Beta function
\item $B$: Reparameterized Beta function
\item $Delta$: Heterozygosity in an individual
\item Y: Hidden Homozygosity state
\item $\Theta$: vector of additional parameters ($\pi$ $N_w$)
\item $Gamma_{wY}$: Posterior probability of Y in w
\item $/eta$: $/alpha + /beta$ for ROH model
\item $c$: Contamination estimate for a library
\item $p_c$: Average proportion of differences between endogenous and contaminating reads
\item $p_e$:Average proportion of differences between endogenous reads
\item $p_a$: Average observed proportion of differences
\item $D_{cor,w}$ Contamination corrected number of differences in $w$
\item $N_{cor,w}$ Contamination corrected number of overlapping sites in $w$
\item $\Xi$: Endogenous read
\item $S_{cor,w}$ $N_{cor,w}$ - $D_{cor,w}$
\item $/mathbf{D_cor}$ Vector of $D_{cor,w}$ 
\item E: Expectation
\item P: Probability
\end{itemize}

@@

We predict the following categories of relatedness with KIn: Identical individuals (twins), First degree (parent-child), First degree (siblings), Second degree, Third degree, Unrelated (including higher than third degree relatives). We show the working of Relatedness HMM with Fig.\ref{fig1:ibd} showing performance of each model, when applied to a pair of siblings.

\begin{figure}[h]
    \centering
    \includegraphics[width=18cm]{plots/plotimg/IBDplot.png}
    \caption{Comparison of IBD fragments predicted by different relatedness models. Top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings. Dashed lines in this panel represent $p_0$, $p_1$ and $p_2$ estimates. The panel just below shows the true IBD state for each window. Each panel below shows the IBD states predicted by a particular relatedness model. The log likelihood value for each model is shown on upper left corner of the panel. }
    \label{fig1:ibd}
\end{figure}

We see that in this case, the siblings model predictions match true IBD states the most. This is reflected in the log likelihoods values (LL) of the models as well, where siblings model has the highest value. We take a loglikelihood ratio ($\Lambda$) of the two most likely models to get a confidence in the relatedness estimate (in this case $\Lambda$ = 22.48). All the models in Fig.\ref{fig1:ibd} utilize information about ROH given by the ROH-HMM.

Performance of ROH-HMM under different cases of ascertainment, contamination and long ROH, and for different coverages is shown in Fig.\ref{fig2:ROH}. We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.1x.

\begin{figure}[h]
    \centering
    \includegraphics[width=18cm]{plots/inkscape_finalImg/ROHplot.png}
    \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment. Top panel shows proportion of differences in a simulated individual along the genome, and the bottom panel shows probability of seeing no ROH. (A) Simulation with no ascertainment, contamination, or ROH. (B) Simulation with ROH. (C) Simulation with ROH and ascertainment. (D) Simulation with ROH and contamination.}
    \label{fig2:ROH}
\end{figure}

We plotted the true positive and false positive rates with a cutoff on the loglikelihood ratio in Fig.\ref{fig3:cutoff}. Here, and in Fig.\ref{fig4:Comparison_READ_KIn}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'. Unrelated label here refers to KIn performance results when all unrelated, degree5, degree4, degree3 pairs are labelled as unrelated. This is useful for comparison to READ, since READ classifies relatedness upto degree2. 'Unrelated w/o degree3'  refers to the performance results when degree 3 is considered as separate from the unrelated individuals. Fig.\ref{fig3:cutoff} shows that for identical, 1st degree, unrelated pairs, the false positive rate is below $5\%$ even when no cut off is used. For second degree, at 0.1x we see a high false positive rate, that does not go down much with an increase in cutoff. For the same relationship, at a coverage of 0.2x, a cutoff of 1 seems appropriate to reduce false positive rate to below $5\%$. Similarly, for siblings at 0.1x, a cutoff of 1 works well to keep false positives low. For parent-child, we find that a cutoff of about 2 is a better choice. Degree 3 shows higher false positive rates in general, and we see that only for 4x coverage and a cutoff of 2, do we see a false positive rate of around $5\%$. In conclusion, it is recommended to use a cutoff of 1 for most cases, with the exception of differentiation between parent-child and siblings, where a cutoff of 2 may be more appropriate, and degree 3 where cutoff of 2 at 4x coverage seems reasonable.

\begin{figure}[h]
    \centering
    \includegraphics[width=18cm]{plots/plotimg/contam0_inbred0_model_performance_allroc_asc0_plot.png}
    \caption{False positive and True Positive rates as a function of cutoff on log likelihood ratio. In top panel, dotted line represents $5\%$}.
    \label{fig3:cutoff}
\end{figure}
